\proglang{collapse}: Advanced and Fast Statistical Computing and Data Transformation in \proglang{R}

### 2024-03-06 misc/JSS article/jss-article-rnw/article.Rnw

%% - \Abstract{} almost as usual
\Abstract{
\pkg{collapse} is a large C/C++-based infrastructure package facilitating complex statistical computing, data transformation, and exploration tasks in \proglang{R} - at outstanding levels of performance and programming efficiency. It also implements a class-agnostic approach to \proglang{R} programming, supporting vector, matrix and data frame-like objects and their popular variants (e.g., \class{factor}, \class{ts}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}), enabling seamless integration with large parts of the \proglang{R} ecosystem. This article introduces the package's key components and design principles in a structured way, supported by a rich set of examples. A small benchmark demonstrates its computational performance.
%  This short article illustrates how to write a manuscript for the
%  \emph{Journal of Statistical Software} (JSS) using its {\LaTeX} style files.
%  Generally, we ask to follow JSS's style guide and FAQs precisely. Also,
%  it is recommended to keep the {\LaTeX} code as simple as possible,
%  i.e., avoid inclusion of packages/commands that are not necessary.
%  For outlining the typical structure of a JSS article some brief text snippets
%  are employed that have been inspired by \cite{Zeileis+Kleiber+Jackman:2008},
%  discussing count data regression in \proglang{R}. Editorial comments and
%  instructions are marked by vertical bars.
}

@@

\pkg{collapse} relates to and integrates key elements from these projects. It offers \pkg{tidyverse}-like data manipulation at the speed and stability of \pkg{data.table} for any data frame-like object. It can turn any vector/matrix/data frame into a time-aware indexed series or frame and perform operations such as lagging, differencing, scaling or centering, encompassing and enhancing core manipulation functionality of \pkg{plm}, \pkg{fixest}, and \pkg{xts}. It also performs fast (grouped, weighted) statistical computations along the columns of matrix-like objects, complementing and enhancing \pkg{matrixStats} and \pkg{Rfast}. Its low-level vectorizations and workhorse algorithms are accessible at the \proglang{R} and C-levels, unlike \pkg{data.table}, where most vectorizations and algorithms are internal. It also supports variable labels and intelligently preserves attributes of all objects, complementing \pkg{labelled}. It provides general (recursive) tools to deal with nested data, enhancing \pkg{tidyr}, \pkg{purrr}, and \pkg{rrapply}. Finally, it provides a small but consistent and powerful set of descriptive statistical tools, yielding sufficient detail for most data exploration purposes, requiring users to invoke packages like \pkg{DescTools} or \pkg{survey} only for specific statistics. In summary, \pkg{collapse} is a foundation package for statistical computing and data manipulation in \proglang{R} that enhances and integrates seamlessly with the \proglang{R} ecosystem while being outstandingly computationally efficient. A significant benefit is that, rather than piecing together a somewhat fragmented ecosystem oriented at different classes and tasks, core computational tasks can be done with \pkg{collapse}, and easily extended by more specialized packages. This tends to result in \proglang{R} scripts that are shorter, more efficient, and more lightweight in dependencies. \newline
% TODO: Need examples of this

@@

\pkg{collapse} is too large and complex to fully present it in a single article, or even to present selected topics in depth. The following sections therefore briefly introduce its key components: (\ref{sec:fast_stat_fun}) the \emph{Fast Statistical Functions} and their (\ref{sec:integration}) integration with data manipulation functions; (\ref{sec:ts_ps}) architecture for time series and panel data; (\ref{sec:join_pivot}) joins and reshaping; (\ref{sec:list_proc}) list processing functions; (\ref{sec:summ_stat}) descriptive tools; and (\ref{sec:glob_opt}) global options. Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes. For deeper engagement with \pkg{collapse}, the short vignette summarizing available \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation and resources} is an excellent starting point.

@@

with arguments \code{x} - data (vector, matrix or data frame-like), \code{g} - groups (atomic vector, list of vectors, or \class{GRP} object), \code{w} - weights, \code{TRA} - transformation, \code{na.rm} - missing values, \code{use.g.names} -  attach group names upon aggregation (if \code{g} is used), \code{drop} - drop dimensions (i.e., simplify to atomic vector if \code{is.null(g)} and \code{x} is matrix or data frame-like), \code{nthreads} - multithreading.\footnote{Not all functions are multithreaded, and parallelism is implemented differently for different functions (detailed in the documentation). The use of single instruction multiple data (SIMD) parallelism in single-threaded mode also implies limited gains from multithreading for simple operations such as \fct{fsum}.} The following examples, taken from the \href{https://sebkrantz.github.io/collapse/articles/collapse_for_tidyverse_users.html#using-the-fast-statistical-functions}{\pkg{collapse} for \pkg{tidyverse} Users} vignette (and executed there) demonstrate their basic usage to calculate (column-wise, grouped, weighted) statistics on different objects. As laid out in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{vignette on object handling}, statistical functions have basis S3 methods for vectors (\class{default}), \class{matrix}, and \class{data.frame}, which call corresponding C implementations that intelligently preserve object attributes. Thus, the functions can be applied to a broad set of \class{matrix} or \class{data.frame}-based objects without the need to define explicit methods. Users can also directly call the basis methods in case S3 dispatch does not yield the intended outcome. For example, \code{fmean.default(EuStockMarkets)} computes the mean of the entire matrix. %, and attributes are preserved as much as possible ()

@@

fmean(mtcars)

@@

fmean(mtcars$mpg, g = mtcars$cyl, TRA = "fill")

@@

\caption{\label{tab:TRA} Arguments to \code{TRA()}.}

@@

\code{TRA()} is called internally in the \emph{Fast Statistical Functions}, the \code{TRA} argument is passed to \code{FUN}. Thus \code{fmean(x, g, w, TRA = "-")} is equivalent to \code{TRA(x, fmean(x, g, w), "-", g)}. The \code{set} argument can also be passed to \emph{Fast Statistical Functions} to toggle transformation by reference. The following examples demonstrate how this design allows flexible ad-hoc transformations using \proglang{R}'s built-in \code{airquality} dataset with daily measurements in New York from May to September 1973.

@@

This imputes \code{Ozone} and \code{Solar.R} by reference using the month median.

@@

airquality |> fmutate(

@@

  ozone_deg = Ozone / Temp,

@@

Whereas the \code{g} argument supports ad-hoc grouping with vectors and lists/data frames, the cost of grouping can be optimized by using factors or, even better, \class{GRP} objects, which readily contain all information \pkg{collapse}'s vectorized statistical functions might require to operate across groups. These objects can be created with \code{GRP()}. The syntax is

@@

 The example below creates and displays a \class{GRP} object from 3 columns in \code{mtcars}. The \code{by} argument also supports column names or indices, and \code{X} could also be an atomic vector.

@@

dat <- get_vars(mtcars, c("mpg", "disp")); w <- mtcars$wt

@@

Similar, data can be transformed, here using the S3 generic \fct{fscale} function.

@@

This programming access can become very useful. For example, the useR 2022 presentation \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{Slides 18-19} aggregates the \href{https://worldmrio.com/}{EORA Global Supply Chain Database} from the country to the world region level. After defining a single grouping object, a list of value-added shares matrices (VB) and outputs for years 1990-2021, is aggregated with no grouping cost using a single line of code like \code{lapply(VB_list, function(x) x$VB |> fsum(g) |> t() |> fmean(g, x$O) |> t())}. On an M1 Mac using 4 threads, this computation, involving 44.7 million summations and 2.6 million weighted means, takes only 0.33 seconds.\footnote{Another example from my recent work involved numerically optimizing a parameter $a$ in an equation of the form $y_j = \sum_i x_{ij}^a\ \forall j\in J$ where there are $J$ groups (1 million in my case), and the optimal value of $a$ is determined by the proximity of the aggregated vector \textbf{y} to another vector \textbf{z}. Thus each iteration of the numerical routine raises the vector \textbf{x} to a different power ($a$), sums it in 1 million groups ($j$) to generate \textbf{y}, and computes the Euclidean distance to \textbf{z} (using \code{collapse::fdist}). Without grouping objects and vectorization, this would have been difficult to handle within reasonable computing times (of a few seconds on the M1).}

@@

\pkg{collapse} also provides a broad set of \href{https://sebkrantz.github.io/collapse/reference/fast-data-manipulation.html}{fast data manipulation functions} familiar to \proglang{R} and \pkg{tidyverse} users, including \fct{fselect}, \fct{fsubset}, \fct{fgroup\_by}, \fct{fsummarise}, \fct{ftransform}, \fct{fmutate}, \fct{across}, \fct{frename}, \fct{fcount}, etc. These are integrated with the \emph{Fast Statistical Functions} to enable vectorized statistical operations in a familiar data frame oriented and \pkg{tidyverse}-like workflow. For example, the following code calculates

@@

mtcars |>
  fsubset(mpg > 11) |>
  fgroup_by(cyl, vs, am) |>

@@

the mean of columns \code{mpg}, \code{carb} and \code{hp}, and the weighted mean of \code{qsec}, after subsetting and grouping the data. This code is very fast (especially with many groups) because data does not need to be split by groups at all. There is also no need to call \code{lapply()} inside the \code{across()} statement: \code{fmean.data.frame()} is applied to a subset of the data containing the three columns.\footnote{Internally, the \code{g} argument of the statistical functions is set as a keyword argument by \code{fsummarise/across} and the function is evaluated on a suitable subset of columns. Thus \code{w} becomes the second positional argument...} The \emph{Fast Statistical Functions} also have a method for grouped data, so \code{fsummarise} is not always needed. The following code calculates weighted group means. By default (\code{keep.w = TRUE}) \code{fmean.grouped\_df} also sums the weights in each group.\footnote{\class{grouped\_df} methods in \pkg{collapse} support grouped data created with either \fct{fgroup\_by} or \fct{dplyr::fgroup\_by}. The latter requires an additional C routine to convert the grouping object to a \class{GRP} object, and is thus much less efficient.}

@@

mtcars |>
  fsubset(mpg > 11, cyl, vs, am, mpg, carb, hp, wt) |>
  fgroup_by(cyl, vs, am) |>
  fmean(wt) |> head(2)

@@

\fct{fsummarise} and \fct{fmutate} can also evaluate other statistical functions in the classical way (split-apply-combine) and handle more complex expressions involving multiple columns and/or functions. However, using any \emph{Fast Statistical Function} causes the whole expression to be vectorized, i.e., evaluated only once and not for every group. This eager vectorization approach is helpful for efficient grouped calculation of more complex statistics. The example below calculates grouped (\code{vs}) bivariate regression slopes (\code{mpg ~ carb}) in a vectorized way.

@@

mtcars |>
 fgroup_by(vs) |>

@@

Apart from vectorization, this code avoids 3 intermediate copies: (1) \code{fmean(carb, TRA = "-")} avoids an expanded vector of group means, (2) \code{fsum(mpg, dm\_carb)} uses the weights (\code{w}) argument to \fct{fsum} to avoid materializing a multiplication (as in \code{fsum(mpg * dm\_carb)}), and (3) division by reference (\code{\%/=\%}) avoids allocating an additional vector for the final result. Under the hood, the expression boils down to an (expensive) grouping step, 5 allocations (of which 2 full length), and 6 loops in C to calculate the result. Any modern laptop can calculate 1 million regression slopes in less than 1 second like this. Another neat example, shared by Andrew Ghazi in a recent \href{https://andrewghazi.github.io/posts/collapse_is_sick/sick.html}{blog post}, vectorizes an expression to compute the p-value across 300k groups for a simulation study, yielding a 70x performance increase. \newline

@@

mtcars |>
    fgroup_by(cyl, vs, am) |>

@@

Both weighted mode and quantiles have a sub-column parallel implementation,\footnote{Use \code{set\_collapse(nthreads = \#)} or the \code{nthreads} arguments to \code{fnth/fmedian/fmode} (default 1).} and, as in this case, can also harness an (optional) optimization by computing an overall ordering vector and passing it to the functions to avoids repeated sorting (using quickselect) of the same elements within each group. For advanced data aggregation, \pkg{collapse} also provides a convenience function, \fct{collap}, which (by default) uses \code{fmean} for numeric and \code{fmode} for non-numeric columns. Below, it aggregates GDP per Capita, Life Expectancy, and country name by World Bank Income Group, with population weights.\footnote{\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}} is a dataset supplied by \pkg{collapse}, extracted from the World Bank World Development Indicators.} This yields population-weighted statistics, the largest country, and each income group's total population (sum of weights) for each year, preserving (default \code{keep.col.order = TRUE}) the order of columns.

@@

\pkg{collapse} also provides a flexible high-performance architecture to perform (\emph{time aware}) computations on time series and panel series. In particular, the user enjoys great flexibility in deciding the desired degree of indexation and mode of computation. It is possible to apply time series and panel data transformations without any indexation by passing individual and/or time identifiers to the respective functions in an ad-hoc fashion, or by using \class{indexed\_frame} and \class{indexes\_series} classes, which implement full and deep indexation. Table~\ref{tab:TSfun} summarizes \pkg{collapse}'s time series and panel data architecture.

@@

\code{findex\_by(), findex(), unindex(), reindex(), timeid(), is\_irregular(), to\_plm()} $+$ rich set of S3 methods for 'indexed\_frame', 'indexed\_series' and 'index\_df'. \\\\

@@

\code{flag(), fdiff(), fgrowth(), fcumsum(), psmat()} [panel data to array conversions] \code{psacf(), pspacf(), psccf()} [autocorrelation functions for panel data] \\\\

@@

\code{f[hd]between(), f[hd]within(), fscale()} [scaling and (higher-dimensional) centering] \\\\

@@

\code{fsubset(), funique(), roworder[v]()} [internal], and \code{na\_omit()} [internal] \\\\
\emph{Summary Functions with Supporting Methods} \\
\code{qsu(), varying()} [panel-variance decomposed statistics] \\

@@

Time series functions such as \fct{fgrowth} (to compute growth rates) are S3 generic and can be applied to most time series classes. In addition to a \code{g} argument for grouped computation, these functions also have a \code{t} argument for indexation. If \code{t} is a plain numeric vector or a factor, it is coerced to integer and interpreted as time steps.\footnote{This is premised on the observation that the most common form of temporal identifier is a numeric variable denoting calendar years.} If \code{t} is a numeric time object (e.g., \class{Date}, \class{POSIXct}, etc.), then it is internally passed through \code{timeid()} which computes the greatest common divisor (GCD) and generates an integer time-id. For the GCD approach to work, \code{t} must have an appropriate class, e.g., for monthly/quarterly data, \code{zoo::yearmon()/zoo::yearqtr()} should be used instead of \class{Date} or \class{POSIXct}. % The following provides a basic example:

@@

data_ix = findex_by(data, id1, ..., time)
data_ix$indexed_series
with(data, indexed_series)
index_df = findex(data_ix)

@@

data = unindex(data_ix)
data_ix = reindex(data, index = index_df)
indexed_series = reindex(vec/mat, index = vec/index_df)

@@

LIFEEXi = wldi$LIFEEX
str(LIFEEXi, width = 70, strict = "cut")

@@

The second option is a wider pivot with \code{how = "wider"}. Here, \code{names} and \code{labels} can be used to select columns containing the names of new columns and their labels.\footnote{multiple columns with names and labels could be selected, which would be combined using \code{"\_"} for names and \code{" - "} for labels.} \emph{Note} how the labels are combined with existing labels such that also this operation is without loss of information. It is, however, a destructive operation, i.e., with 2 or more columns selected through \code{values}, \fct{pivot} is not able to reverse it. Further arguments like \code{na.rm}, \code{fill}, \code{sort}, and \code{transpose} can be used to control the casting process.
%
<<pivot_wider>>=
(dw <- pivot(data, "id", names = "type", labels = "type_name", how = "w"))
namlab(dw)
@
%
For the recast pivot (\code{how = "recast"}), unless a column named \code{variable} exists in the data, the source and (optionally) destination of variable names need to be specified using a list passed to \code{names}, and similarly for \code{labels}. Again, taking along labels is optional, and omitting either the list's \code{from} or \code{to} elements will omit the respective operations.
%
<<pivot_recast>>=
(dr <- pivot(data, ids = "id", names = list(from = "type"),
             labels = list(from = "type_name", to = "label"), how = "r"))
vlabels(dr)
@
%
As with the other pivots, this operation does not incur any loss of information. A suitable reverse operation also exists: \code{pivot(dr, "id", names = list(to = "type"), labels = list(from = "label", to = "type_name"), how = "r")}. More features of \fct{pivot} are demonstrated in the \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples}. Notably, it is also possible to perform longer and recast pivots without id variables. The recast pivot without ids resembles a generalization of \code{data.table::transpose()}, albeit slightly less efficient.
%
\section{List processing} \label{sec:list_proc}
%
Often in programming, nested structures are needed. A typical use case involves running statistical procedures for multiple configurations of variables and parameters and saving multiple objects (such as a model object, performance statistics, and predictions) in a list. Nested data is also often the result of web scraping or web APIs. A typical use case in development involves serving different data according to user choices, e.g., in response to nested user inputs in shiny apps. Except for certain recursive functions found in packages such as \pkg{purr}, \pkg{tidyr}, or \pkg{rrapply}, \proglang{R} lacks a general recursive toolkit to create, query, and tidy nested data. \pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/list-processing.html}{list processing functions} attempt to provide a basic toolkit. \newline

>>>>

%% - \Abstract{} almost as usual
\Abstract{
\pkg{collapse} is a large C/C++-based infrastructure package facilitating complex statistical computing, data transformation, and exploration tasks in \proglang{R} - at outstanding levels of performance and programming efficiency. It also implements a class-agnostic approach to \proglang{R} programming, supporting vector, matrix and data frame-like objects and their popular variants (e.g., \class{factor}, \class{ts}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}), enabling its seamless integration with large parts of the \proglang{R} ecosystem. This article introduces the package's key components and design principles in a structured way, supported by a rich set of examples. A small benchmark demonstrates its computational performance.
%  This short article illustrates how to write a manuscript for the
%  \emph{Journal of Statistical Software} (JSS) using its {\LaTeX} style files.
%  Generally, we ask to follow JSS's style guide and FAQs precisely. Also,
%  it is recommended to keep the {\LaTeX} code as simple as possible,
%  i.e., avoid inclusion of packages/commands that are not necessary.
%  For outlining the typical structure of a JSS article some brief text snippets
%  are employed that have been inspired by \cite{Zeileis+Kleiber+Jackman:2008},
%  discussing count data regression in \proglang{R}. Editorial comments and
%  instructions are marked by vertical bars.
}

@@

\pkg{collapse} relates to and integrates key elements from these projects. It offers \pkg{tidyverse}-like data manipulation at the speed and stability of \pkg{data.table} for any data frame-like object. It can turn any vector/matrix/data frame into a time-aware indexed series or frame and perform operations such as lagging, differencing, scaling or centering, encompassing and enhancing core manipulation functionality of \pkg{plm}, \pkg{fixest}, and \pkg{xts}. It also performs fast (grouped, weighted) statistical computations along the columns of matrix-like objects, complementing and enhancing \pkg{matrixStats} and \pkg{Rfast}. Its low-level vectorizations and workhorse algorithms are accessible at the \proglang{R} and C-levels, unlike \pkg{data.table}, where most vectorizations and algorithms are internal. It also supports variable labels and intelligently preserves attributes of all objects, complementing \pkg{labelled}. It provides general (recursive) tools to deal with nested data, enhancing \pkg{tidyr}, \pkg{purrr}, and \pkg{rrapply}. Finally, it provides a small but consistent and powerful set of descriptive statistical tools, yielding sufficient detail for most data exploration purposes, requiring users to invoke packages like \pkg{DescTools} or \pkg{survey} only for specific statistics. In summary, \pkg{collapse} is a foundation package for statistical computing and data manipulation in \proglang{R} that enhances and integrates seamlessly with the \proglang{R} ecosystem while being outstandingly computationally efficient. A significant benefit is that, rather than piecing together a fragmented ecosystem oriented at different classes and tasks, many core computational tasks can be done with \pkg{collapse}, and easily extended by more specialized packages. This tends to result in \proglang{R} scripts that are shorter, more efficient, and more lightweight in dependencies. \newline
% TODO: Need examples of this

@@

\pkg{collapse} is too large and complex to fully present it in a single article, or even to present selected topics in depth. The following sections therefore briefly introduce its key components: (\ref{sec:fast_stat_fun}) the \emph{Fast Statistical Functions} and their (\ref{sec:integration}) integration with data manipulation functions; (\ref{sec:ts_ps}) architecture for time series and panel data; (\ref{sec:join_pivot}) joins and reshaping; (\ref{sec:list_proc}) list processing functions; (\ref{sec:summ_stat}) descriptive tools; and (\ref{sec:glob_opt}) global options. Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes. For deeper engagement with \pkg{collapse}, a short vignette summarizing available \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation and resources} is an excellent starting point.

@@

with arguments \code{x} - data (vector, matrix or data frame-like), \code{g} - groups (atomic vector, list of vectors, or \class{GRP} object), \code{w} - weights, \code{TRA} - transformation, \code{na.rm} - missing values, \code{use.g.names} -  attach group names upon aggregation (if \code{g} is used), \code{drop} - drop dimensions (i.e., simplify to atomic vector if \code{is.null(g)} and \code{x} is matrix or data frame-like), \code{nthreads} - multithreading.\footnote{Not all functions are multithreaded, and parallelism is implemented differently for different functions (detailed in the documentation). The use of single instruction multiple data (SIMD) parallelism in single-threaded mode also implies limited gains from multithreading for simple operations such as \fct{fsum}.} The following examples, taken from the \href{https://sebkrantz.github.io/collapse/articles/collapse_for_tidyverse_users.html#using-the-fast-statistical-functions}{\pkg{collapse} for \pkg{tidyverse} Users} vignette demonstrate their basic usage to calculate (column-wise, grouped, weighted) statistics on different objects. As laid out in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{vignette on object handling}, statistical functions have basis S3 methods for vectors (\class{default}), \class{matrix}, and \class{data.frame}, which call corresponding C implementations that intelligently preserve object attributes. Thus, the functions can be applied to a broad set of \class{matrix} or \class{data.frame}-based objects without the need to define explicit methods. Users can also directly call the basis methods in case S3 dispatch does not yield the intended outcome. For example, \code{fmean.default(EuStockMarkets)} computes the mean of the entire matrix. %, and attributes are preserved as much as possible ()
<<faststatfun>>=

@@

fmean(mtcars[5:10])

@@

fmean(mtcars$mpg, g = mtcars$cyl, TRA = "fill") |> head(20)

@@

\caption{\label{tab:TRA} Available \code{FUN} choices in \code{TRA()}.}

@@

\code{TRA()} is called internally in the \emph{Fast Statistical Functions}, the \code{TRA} argument is passed to \code{FUN}. Thus \code{fmean(x, g, w, TRA = "-")} is equivalent to \code{TRA(x, fmean(x, g, w), "-", g)}. The \code{set} argument can also be passed to \emph{Fast Statistical Functions} to toggle transformation by reference. The following examples demonstrate how this design allows flexible ad-hoc transformations using \proglang{R}'s built-in \href{https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/airquality}{\code{airquality} dataset} with daily measurements in New York from May to September 1973.

@@

This imputes columns \code{Ozone} and \code{Solar.R} by reference using the month median.

@@

airquality |> fmutate(ozone_deg = Ozone / Temp,

@@

@@

Whereas the \code{g} argument supports ad-hoc grouping with vectors and lists/data frames, the cost of grouping can be optimized by using factors or, even better, \class{GRP} objects, which readily contain all information \pkg{collapse}'s vectorized statistical functions might require to operate across groups. These objects can be created with \code{GRP()}. Its syntax is

@@

 The example below creates and displays a \class{GRP} object from 3 columns in \href{https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/mtcars}{\code{mtcars}}. The \code{by} argument also supports column names or indices, and \code{X} could also be an atomic vector.

@@

dat <- get_vars(mtcars, c("mpg", "disp")); w <- mtcars$wt;

@@

Similarly, data can be transformed, here using the S3 generic \fct{fscale} function.

@@

This programming access can become very useful. For example, the useR 2022 presentation \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{Slides 18-19} aggregates the \href{https://worldmrio.com/}{EORA Global Supply Chain Database} from the country to the world region level. After defining a single grouping object, a list of value-added shares matrices (VB) and outputs (O) for years 1990-2021, is aggregated with no grouping cost using a single line of code like \code{lapply(VB_list, function(x) x$VB |> fsum(g) |> t() |> fmean(g, x$O) |> t())}. On an M1 Mac using 4 threads, this computation, involving 44.7 million summations and 2.6 million weighted means, takes only 0.33 seconds.\footnote{Another recent example involved numerically optimizing a parameter $a$ in an equation of the form $y_j = \sum_i x_{ij}^a\ \forall j\in J$ where there are $J$ groups (1 million in my case), and the optimal value of $a$ is determined by the proximity of the aggregated vector \textbf{y} to another vector \textbf{z}. Thus each iteration of the numerical routine raises the vector \textbf{x} to a different power ($a$), sums it in 1 million groups ($j$) to generate \textbf{y}, and computes the Euclidean distance to \textbf{z} (using \code{collapse::fdist}). Without grouping objects and vectorization, this would have been difficult to handle within reasonable computing times (of a few seconds on the M1).}

@@

\pkg{collapse} also provides a broad set of \href{https://sebkrantz.github.io/collapse/reference/fast-data-manipulation.html}{fast data manipulation functions} familiar to \proglang{R} and \pkg{tidyverse} users, including \fct{fselect}, \fct{fsubset}, \fct{fgroup\_by}, \fct{fsummarise}, \fct{ftransform}, \fct{fmutate}, \fct{across}, \fct{frename}, \fct{fcount}, etc. These are integrated with the \emph{Fast Statistical Functions} to enable vectorized statistical operations in a familiar data frame oriented and \pkg{tidyverse}-like workflow. For example, the following code calculates the mean of columns

@@

mtcars |> fsubset(mpg > 11) |> fgroup_by(cyl, vs, am) |>

@@

 \code{mpg}, \code{carb} and \code{hp}, and the weighted mean of \code{qsec}, after subsetting and grouping the data. This code is very fast (especially with many groups) because data does not need to be split by groups at all. There is also no need to call \code{lapply()} inside the \code{across()} statement: \code{fmean.data.frame()} is applied to a subset of the data containing the three columns.\footnote{Internally, the \code{g} argument of the statistical functions is set as a keyword argument by \code{fsummarise/across} and the function is evaluated on a suitable subset of columns. Thus \code{w} becomes the second positional argument...} The \emph{Fast Statistical Functions} also have a method for grouped data, so \code{fsummarise} is not always needed. The following example calculates weighted group means. By default (\code{keep.w = TRUE}) \code{fmean.grouped\_df} also sums the weights in each group.\footnote{\class{grouped\_df} methods in \pkg{collapse} support grouped data created with either \fct{fgroup\_by} or \fct{dplyr::group\_by}. The latter requires an additional C routine to convert the \pkg{dplyr} grouping object to a \class{GRP} object, and is thus less efficient.}

@@

mtcars |> fsubset(mpg > 11, cyl, vs, am, mpg, carb, hp, wt) |>
  fgroup_by(cyl, vs, am) |> fmean(wt) |> head(2)

@@

\fct{fsummarise} and \fct{fmutate} can also evaluate arbitrary statistical functions in the classical way (split-apply-combine) and handle more complex expressions involving multiple columns and/or functions. However, using any \emph{Fast Statistical Function} causes the whole expression to be vectorized, i.e., evaluated only once and not for every group. This eager vectorization approach enables efficient grouped calculation of more complex statistics. The example below calculates grouped (\code{vs}) bivariate regression slopes (\code{mpg ~ carb}) in a vectorized way.

@@

mtcars |> fgroup_by(vs) |>

@@

Apart from vectorization, this code avoids 3 intermediate copies: (1) \code{fmean(carb, TRA = "-")} avoids an expanded vector of group means, (2) \code{fsum(mpg, dm\_carb)} uses the weights (\code{w}) argument to \fct{fsum} to avoid materializing a multiplication (as in \code{fsum(mpg * dm\_carb)}), and (3) division by reference (\code{\%/=\%}) avoids allocating an additional vector for the final result. Under the hood, the expression boils down to an (expensive) grouping step, 5 allocations (of which 2 full length), and 6 loops in C to calculate the result. Any modern laptop can calculate 1 million regression slopes in less than 1 second like this. Another very neat example, shared by Andrew Ghazi in a recent \href{https://andrewghazi.github.io/posts/collapse\_is\_sick/sick.html}{blog post},\footnote{https://andrewghazi.github.io/posts/collapse\_is\_sick/sick.html} vectorizes an expression to compute the $p$~value across 300k groups for a simulation study, yielding a 70x performance increase over \pkg{dplyr}. \newline

@@

mtcars |> fgroup_by(cyl, vs, am) |>

@@

Both weighted mode and quantiles have a sub-column parallel implementation,\footnote{Use \code{set\_collapse(nthreads = \#)} or the \code{nthreads} arguments to \code{fnth/fmedian/fmode} (default 1).} and, as shown above, can also harness an (optional) optimization by computing an overall ordering vector and passing it to each quantile function to avoid repeated partial sorting (using quickselect) of the same elements within groups. For advanced data aggregation, \pkg{collapse} also provides a convenience function, \fct{collap}, which (by default) uses \code{fmean} for numeric and \code{fmode} for non-numeric columns. Below, it aggregates GDP per capita, life expectancy, and country name by World Bank income group, with population weights.\footnote{\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}} is a dataset supplied by \pkg{collapse}, extracted from the World Bank World Development Indicators.} This yields population-weighted statistics, the largest country, and each income group's total population (sum of weights) for each year, preserving (default \code{keep.col.order = TRUE}) the order of columns.

@@

\pkg{collapse} also provides a flexible high-performance architecture to perform (time aware) computations on time series and panel series. In particular, the user enjoys great flexibility in deciding the desired degree of indexation and mode of computation. It is possible to apply time series and panel data transformations without any indexation by passing individual and/or time identifiers to the respective functions in an ad-hoc fashion, or by using \class{indexed\_frame} and \class{indexes\_series} classes, which implement full and deep indexation. Table~\ref{tab:TSfun} summarizes \pkg{collapse}'s time series and panel data architecture.

@@

\code{findex\_by(), findex(), unindex(), reindex(), timeid(), is\_irregular(), to\_plm()} $+$ S3 methods for \class{indexed\_frame}, \class{indexed\_series} and \class{index\_df} \\\\

@@

\code{flag(), fdiff(), fgrowth(), fcumsum(), psmat()} \\ \code{psacf(), pspacf(), psccf()} \\\\

@@

\code{fscale(), f[hd]between(), f[hd]within()} \\\\

@@

\code{fsubset(), funique(), roworder[v]()} (internal), \code{na\_omit()} (internal) \\\\
\emph{Summary functions with supporting methods} \\
\code{varying(), qsu()} \\

@@

Time series functions such as \fct{fgrowth} (to compute growth rates) are S3 generic and can be applied to most time series classes. In addition to a \code{g} argument for grouped computation, these functions also have a \code{t} argument for indexation. If \code{t} is a plain numeric vector or a factor, it is coerced to integer and interpreted as time steps.\footnote{This is premised on the observation that the most common form of temporal identifier is a numeric variable denoting calendar years. Users need to manually call \code{timeid()} on plain numeric vectors with decimals to yield an appropriate integer representation.} If \code{t} is a numeric time object (e.g., \class{Date}, \class{POSIXct}, etc.), then it is internally passed through \code{timeid()} which computes the greatest common divisor (GCD) and generates an integer time-id. For the GCD approach to work, \code{t} must have an appropriate class, e.g., for monthly/quarterly data, \code{zoo::yearmon()/zoo::yearqtr()} should be used instead of \class{Date} or \class{POSIXct}. % The following provides a basic example:

@@

data_ix <- findex_by(data, id1, ..., time)
data_ix$indexed_series; with(data, indexed_series)
index_df <- findex(data_ix)

@@

data <- unindex(data_ix)
data_ix <- reindex(data, index = index_df)
indexed_series <- reindex(vec/mat, index = vec/index_df)

@@

LIFEEXi <- wldi$LIFEEX; str(LIFEEXi, width = 70, strict = "cut")

@@

The second option is a wider pivot with \code{how = "wider"}. Here, \code{names} and \code{labels} can be used to select columns containing the names of new columns and their labels.\footnote{multiple columns with names and labels could be selected, which would be combined using \code{"\_"} for names and \code{" - "} for labels.} Note how the labels are combined with existing labels such that also this operation is without loss of information. It is, however, a destructive operation, i.e., with 2 or more columns selected through \code{values}, \fct{pivot} is not able to reverse it. Further arguments like \code{na.rm}, \code{fill}, \code{sort}, and \code{transpose} can be used to control the casting process.
%
<<pivot_wider>>=
(dw <- pivot(data, "id", names = "type", labels = "type_name", how = "w"))
namlab(dw)
@
%
For the recast pivot (\code{how = "recast"}), unless a column named \code{variable} exists in the data, the source and (optionally) destination of variable names need to be specified using a list passed to \code{names}, and similarly for \code{labels}. Again, taking along labels is optional, and omitting either the list's \code{from} or \code{to} elements will omit the respective operations.
%
<<pivot_recast>>=
(dr <- pivot(data, ids = "id", names = list(from = "type"),
             labels = list(from = "type_name", to = "label"), how = "r"))
vlabels(dr)
@
%
As with the other pivots, this operation does not incur any loss of information. A suitable reverse operation also exists: \code{pivot(dr, "id", names = list(to = "type"), labels = list(from = "label", to = "type_name"), how = "r")}. More features of \fct{pivot} are demonstrated in the \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples}. Notably, it is also possible to perform longer and recast pivots without id variables. The recast pivot without ids resembles a generalization of \code{data.table::transpose()}, albeit slightly less efficient.
%
\section{List processing} \label{sec:list_proc}
%
Often in programming, nested structures are needed. A typical use case involves running statistical procedures for multiple configurations of variables and parameters and saving multiple objects (such as a model object, performance statistics, and predictions) in a list. Nested data is also often the result of web scraping or web APIs. A typical use case in development involves serving different data according to user choices, e.g., in response to nested user inputs in shiny apps. Except for certain recursive functions found in packages such as \pkg{purr}, \pkg{tidyr}, or \pkg{rrapply}, \proglang{R} lacks a general recursive toolkit to create, query, and tidy nested data. \pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/list-processing.html}{list processing functions} attempt to provide a basic toolkit. \newline

To create nested data, \fct{rsplit} generalizes \fct{split} and (recursively) splits up data frame-like objects into a (nested) list.
<<list_proc>>=
(dl <- mtcars |> rsplit(mpg + hp + carb ~ vs + am)) |> str(max.level = 2)
